\PaperTitle

### 2020-05-22 paper/filter_bubbles.tex
shorthand

\documentclass[format=acmsmall, review=false]{acmart}
\settopmatter{printacmref=false, printccs=false}
\usepackage{booktabs} % For formal tables
\usepackage[ruled]{algorithm2e} % For algorithms
\usepackage{enumitem}
\usepackage{subcaption}
\newcommand{\xhdr}[1]{\vspace{1mm} \noindent{\bf #1}}
\renewcommand{\footnotetextcopyrightpermission}[1]{} % remove copyright block
\renewcommand{\footnotetextauthorsaddresses}[1]{} % remove author address 
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

@@

% Title. Note the optional short title for running heads. In the interest of anonymization, please do not include any acknowledgements.
\newcommand\PaperTitle{Deconstructing the Filter Bubble: \\ Consumer Decision-Making and Recommender Systems}
\title[\PaperTitle]{\PaperTitle}

@@

% Abstract. Note that this must come before \maketitle.
\begin{abstract}
We study a model of user decision-making in the context of recommender systems via numerical simulation. Our model provides an explanation for the findings of Nguyen, et. al (2014), where users consume increasingly similar items over time even without recommendation. Our model naturally generates such filter bubble effects since users face a sequential decision making problem under uncertainty and there are informational spillovers where a user consuming a item and learning how she values it reduces uncertainty and provides information about similar items. As shown in Nguyen, et. al (2014), recommendation alleviates these filter bubble effects and we further find that, by providing users with information, recommendation leads users to consume increasingly similar items. Finally, our model highlights the importance of collecting data on user beliefs and their evolution over time both to design better recommendations and to further understand their impact.
\end{abstract}

@@

Recommender systems (RS) have become critical for assisting users in navigating the large choice sets that they face on many online platforms. For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube. Users in many cases are not aware of most items, let alone have full information about their preferences over them. To make matters worse, the item in these contexts are usually experience goods whose true value can only be learned after consumption.
\par

Recommender systems have driven a significant fraction of consumer choice on these platforms with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al. (2013, Oct.),  How retailers can keep up with consumers. \url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}. Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems. There have been claims that personalized recommender systems lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized recommender systems may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
\par
Understanding how recommender systems influence user behavior is important not only for characterizing the broader consequences of such systems but also for guiding their design. In this paper we develop a theoretical model of user decision-making in contexts where recommender systems are traditionally deployed. We utilize previous empirical studies that characterize how recommender systems influence user choice as a benchmark and our theoretical model provides an intuitive mechanism that can explain these empirical results. The key insight of our model is that user \textit{beliefs} drive the consumption choices of users and that recommendations provide them with information that leads them to update their beliefs and alter their choices. A crucial component of our model is that users' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items. We use these insights to provide guidance for recommender system design. In order to design good recommendations it becomes important for designers to understand user beliefs of the goods they haven't consumed thus far and track how these beliefs evolve over time.

\xhdr{Our Model} We consider a model of user choice that has the following components.
\par
The first component of our model is that users sequentially consume items and face large choice sets. In our setting of interest, users are long-lived, but they only consume a small fraction of this choice set over their lifetime. This is traditionally the case on online platforms that have thousands or millions of options for users.
\par
The second component is that users have \textit{uncertainty} about their valuation of the different items before they consume them. This is motivated both by the fact that many contexts where recommender systems are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why recommender systems exist in the first place. Thus, users face a sequential decision-making problem under uncertainty.
\par 
The third, and most crucial, element is that consumption of an item reveals information that changes user beliefs about the valuation of similar items. Unlike in standard sequential decision-making problems, once an item is consumed all uncertainty about its valuation is resolved and provides information that enables users to update their beliefs about similar items. This exploits the fact that the valuations of similar items are correlated which assists users in navigating the vast product space. The idea that users make similarity-based assessments to guide their choice has grounding in theoretical work in decision theory \cite{gilboa1995case} and empirical evidence of how users navigate large choice sets \cite{schulz2019structured}.
\par 
Finally, in our model recommendation provides users with information about the true valuations. We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component. This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other users' preferences and the common-value component is what the recommender can learn from previous users' data. We suppose that the recommender knows the common-value component for each item and combines it with users' beliefs over the product space when designing personalized recommendation.

@@

Lastly, we discuss how our model and findings can be used to inform the design and evaluation of recommender systems as well as the data that is traditionally collected for them. This highlights the importance of user beliefs in determining user consumption choices and how both recommendation and informational spillovers determine how these beliefs change over time. By collecting information on user beliefs, recommender system designers can understand what items a user would consume \textit{without} recommendation and then predict how providing information to the user would change her beliefs and resulting consumption decisions. Thus,  our evaluation measure determines the value of a recommendation based on the marginal welfare gain associated with providing a user with a recommendation over what the user would do without it. We discuss how this provides an additional rationale as to why ``accurate" recommendations are not always good recommendations.

@@

The first set of related works studies the extent and implications of filter bubbles. \cite{pariser2011filter} first informally described the idea of the ``filter bubble'' which is that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content. Following this a number of empirical studies, done in various disciplines, have since studied the extent to which this phenomenon exists in a wide range of contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}. The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists in the context of movie consumption. They find that even users whose consumption choices are not guided by recommendations exhibit behavior consistent with ``filter bubbles'' and that recommender systems can actually increase the diversity of the content that users consume. To our knowledge there are no theoretical models that rationalize these empirical findings and we provide a theoretical framework through which to view this problem. Furthermore, our paper provides a clear mechanism at play that drives such effects and how recommendation interacts with them.

@@

Another set of papers has examined whether recommender systems can lead users to become increasingly homogenized or consume similar content to one another. \cite{celma2008hits, treviranus2009value} show that incorporating content popularity into recommender systems can lead to increased user homogenization. \cite{chaney2018algorithmic} shows how user homogenization may arise from training recommender systems on data from users exposed to algorithmic recommendations. \cite{fleder2009blockbuster} show that homogenization can increase due to a recommendation bias towards popularity that arises from lack of information about items with limited consumption histories. Our model leads to similar results as previous work where recommender systems lead to increased user homogenization. However, the mechanisms behind this differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.

@@

Another set of papers studies the impact of human decision-making on the design and evaluation of recommender systems. \cite{chen2013human} surveys the literature on the relationship between human decision making and recommender systems. The closest set of papers pointed out in this survey are those related to preference construction \cite{bettman1998constructive, lichtenstein2006construction} whereby users develop preferences over time through the context of a decision process. We point out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where recommender systems are deployed, they have incomplete information of their valuations and both consumption and recommendation provide them with information to reduce their uncertainty. Thus, the primary insight of our paper is that user beliefs and how users update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of recommender systems. Within this literature, \cite{celma2008new, cremonesi2013user, pu2011user} focus on ``user-centric" approaches to recommendation whereby user evaluation of the usefulness of recommendation is a key evaluation measure. Our evaluation measure is similar, but, unlike previous approaches, emphasizes the importance of user beliefs. Finally, \cite{hodgson2019horse} considers a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.

@@

\xhdr{User Learning}
When a user consumes an item $n$ she learns the realized value for that item. We assume that learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying an item provides more information about similar items than about dissimilar ones.\footnote{\cite{schulz2019structured} empirically studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders. They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular good provides them with information about similar items. We incorporate this finding into our model in a stylized manner.} In order to have a well-defined notion of similarity we need to define a distance function between items, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are indices of items in $\mathcal{J}$. This distance function is not intended to model the intricacies of a realistic product space, but instead to provide a stylized product space to help us understand the effects of informational spillovers on user behavior. The basic intuition, in the context of movie consumption, is that a user watching \textit{John Wick} (item $n$) provides more information about how much she will like \textit{John Wick: Chapter Two} (item $m$) than \textit{Titanic} (item $q$) since $d(n, m) < d(n, q)$. 
\par
We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$. The scalar $\rho \in [0,1]$ therefore impacts the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby. Informativeness, for any $\rho \in (0,1)$, is decreasing in distance. The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc. \footnote{This exponential decay correlation structure can be related to the tenet of case-based similarity of \cite{gilboa1995case} -- see \cite{billot2008axiomatization} for an axiomatization of exponential similarity.}
\par
The precise updating rule is as follows. Recall that at time $t$ the user's consumption history is given by $C_{i}^{t}$ and we denote the utility realizations of these items as $c_t$. We denote $\mu_t$ as the initial mean beliefs the user has over the items in $C_{i}^{t}$ and $\mu_{N-t}$ as the initial mean beliefs the user has over the remaining $N-t$ items, $\mathcal{J} \setminus C_{i}^{t}$. We partition the covariance matrix as follows:
\[ \Sigma =  \left( \begin{array}{cc}
\Sigma_{(N-t, N-t)} & \Sigma_{(N-t,t)} \\
\Sigma_{(t,N-t)} & \Sigma_{(t,t)}
\end{array} \right)
\]
Thus, after consuming the items in $C_{i}^{t}$ the resulting beliefs over the remaining items are given by $\mathcal{N}(\bar{\mu}, \bar{\Sigma})$ where $\bar{\mu}$ and $\bar{\Sigma}$ are as follows:
\begin{align*}
\bar{\mu} \mid c_t = \mu_{N-t} + \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1}(c_t - \mu_t) \\
\bar{\Sigma} \mid c_t = \Sigma_{(N-t,N-t)} - \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1} \Sigma_{(t,N-t)}
\end{align*}

@@

\xhdr{Recommendation.}
Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how recommender systems are implemented in practice. We model recommendation as giving users information about the valuation of the products.
\par

@@

We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, \ standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$. The range of considered parameter values cover the relevant portions of the parameter space in order to provide full insight into the behavior of the model. 
\par
When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest. We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.\footnote{In the supplementary material we include the same exercises presented here for $N = 100$ and $N = 500$ and show that the results are qualitatively similar to those reported here.}
\par

\begin{figure}
\caption{Local Consumption and Correlation, $N = 200$}
\begin{subfigure}{.5\linewidth}
  \centering
  \includegraphics[width=.9\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}
  \label{fig:sfig1}
\end{subfigure}%
\begin{subfigure}{.5\linewidth}
  \centering
  \includegraphics[width=.9\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}
  \label{fig:sfig2}
\end{subfigure}
\caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime. The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.}
\label{fig:correlation_consumption_path}
\end{figure}

\section{Results}
\subsection{Local Consumption and Filter Bubbles}
We characterize ``filter bubble'' effects in our model as the degree to which users engage in \textit{local consumption}. We define local consumption in terms of the average consumption distance between the items consumed by the users at time $t-1$ and $t$. Thus, in the context of our model, filter bubble effects arise when the average consumption distance decreases over time and, across regimes, when the overall levels are lower for a given recommendation regime compared to another.

@@

\subsection{User Welfare and Product Diversity}
In this section we primarily focus on the impact of recommendation on user welfare and the overall diversity of the items that they consume. While in the previous section we looked at the distance between consecutive items, in this section we focus on a diversity measure that considers the entire consumed set of items. The diversity measure we utilize is common in recommender system literature (e.g. \cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed items:
$$D_i:=\frac{1}{N}\frac{1}{T(T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$$

\begin{finding}\label{finding_diversity}
The impact of recommendation on product diversity:
\begin{enumerate}
\item When $\rho = 0$, product diversity is the same across all three recommendation regimes;
\item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime. This effect is amplified as $\rho$ increases as well as when users become more risk-averse.
\end{enumerate}
\end{finding}
\par 
Finding~\ref{finding_diversity} summarizes the main results on product diversity. As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes. The over-exploitation of information spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime. As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases. In the supplementary material (Figure 5), we show how diversity varies as $\rho$ increases across the three regimes that we consider. There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in the supplementary material (Figure 9). The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
\par 
We now study how recommendation impacts user welfare in our model. In our model users make consumption decisions that maximize their current period ex-ante utility that depends on their beliefs in that period. Thus, from an ex-ante perspective, they make optimal decisions, but our primary interest is in understanding how the ex-post, or realized, utility varies across regimes and parameter values. We define user's \textit{ex-post} welfare as the average of the realized values, controlling for the effect of $T$:
$$W_i:= \frac{1}{T}\sum_{n \in C_i^T} x_{i,n}$$

>>>>

\documentclass[manuscript]{acmart}
\settopmatter{printacmref=false, printccs=false}
\usepackage{enumitem}
\usepackage{subcaption}
\newcommand{\xhdr}[1]{\vspace{1mm} \noindent{\bf #1}}

@@

% Title. Note the optional short title for running heads. In the interest of anonymization, please do not include any acknowledgements.
\newcommand\PaperTitle{Deconstructing the Filter Bubble: Consumer Decision-Making and RSs}
\title[\PaperTitle]{\PaperTitle}

@@

% Abstract. Note that this must come before \maketitle.
\begin{abstract}
We study a model of user decision-making in the context of RSs via numerical simulation. Our model provides an explanation for the findings of Nguyen, et. al (2014), where users consume increasingly similar items over time even without recommendation. Our model naturally generates such filter bubble effects since users face a sequential decision making problem under uncertainty and there are informational spillovers where a user consuming an item and learning how she values it reduces uncertainty and provides information about similar items.
\par
In line with Nguyen, et. al (2014), recommendation alleviates these filter bubble effects. However, we find that recommendation leads to an increase in homogeneity across users, resulting in a trade-off between homogenizing across user consumption and diversifying within user consumption. Finally, our model highlights the importance of collecting data on user beliefs and their evolution over time both to design better recommendations and to further understand their impact.
\end{abstract}

@@

Recommender systems (RS) have become critical for assisting users in navigating the large choice sets that they face on many online platforms. For instance, users have to select from thousands of movies on Netflix, millions of products on Amazon, and billions of videos on YouTube. Users in many cases are not aware of most items, let alone have full information about their preferences over them. To make matters worse, the items in these contexts are usually experience goods whose true value for users can only be learned after consumption.
\par

RS have driven a significant fraction of consumer choice on these platforms with 75\% of movies watched on Netflix and 35\% of page-views on Amazon coming from recommendations.\footnote{MacKenzie et al. (2013, Oct.),  How retailers can keep up with consumers. \url{https://www.mckinsey.com/industries/retail/our-insights/how-retailers-can-keep-up-with-consumers}. Retrieved on October 3, 2019.} While there are many positive effects from these systems, there is an increasing worry that there are unintended side-effects of recommendation systems. There have been claims that personalized RSs lead users into \textit{filter bubbles} where they effectively get isolated from a diversity of viewpoints or content \cite{pariser2011filter}, and that personalized RSs may also lead users to become increasingly homogenized at the same time \cite{chaney2018algorithmic, hosanagar2013will}.
\par
Understanding how RS influence user behavior is important not only for characterizing the broader consequences of such systems but also for guiding their design. In this paper we develop a theoretical model of user decision-making in contexts where RS are traditionally deployed. We utilize previous empirical studies that characterize how RS influence user choice as a benchmark and our theoretical model provides an intuitive mechanism that can explain these empirical results. The key insight of our model is that user \textit{beliefs} drive the consumption choices of users and that recommendations provide them with information that leads them to update their beliefs and alter their choices. A crucial component of our model is that users' beliefs about items are driven not only by recommendations, but also from their previous experiences with similar items. We use these insights to provide guidance for RS design. In order to design good recommendations it becomes important for designers to understand user beliefs of the goods they haven't consumed thus far and track how these beliefs evolve over time.

\xhdr{Our Model} We consider a model of user choice that has the following components.
\par
The first component of our model is that users sequentially consume items and face large choice sets. In our setting of interest, users are long-lived, but they only consume a small fraction of this choice set over their lifetime. This is traditionally the case on online platforms that have thousands or millions of options for users.
\par
The second component is that users have \textit{uncertainty} about their valuation of the different items before they consume them. This is motivated both by the fact that many contexts where RSs are deployed contain experience goods, whose true value can only be learned after consumption, and the fact that such uncertainty is why RSs exist in the first place. Thus, users face a sequential decision-making problem under uncertainty.
\par 
The third, and most crucial, element is that consumption of an item reveals information that changes user beliefs about the valuation of similar items. Unlike in standard sequential decision-making problems, once an item is consumed all uncertainty about its valuation is resolved and provides information that enables users to update their beliefs about similar items. This exploits the fact that the valuations of similar items are correlated which assists users in navigating the vast product space. The idea that users make similarity-based assessments to guide their choice has grounding in theoretical work in decision theory \cite{gilboa1995case} and empirical evidence of how users navigate large choice sets \cite{schulz2019structured}.
\par 
Finally, in our model recommendation provides users with information about the true valuations. We model the realized valuations as being a weighted sum of a common-value and an idiosyncratic component. This formulation gives a stylized notion of predictability of user preferences where the idiosyncratic component is inherently unpredictable given other users' preferences and the common-value component is what the recommender can learn from previous users' data. We suppose that the recommender knows the common-value component for each item and combines it with users' beliefs over the product space when designing personalized recommendation.

@@

Lastly, we discuss how our model and findings can be used to inform the design and evaluation of RSs as well as the data that is traditionally collected for them. This highlights the importance of user beliefs in determining user consumption choices and how both recommendation and informational spillovers determine how these beliefs change over time. By collecting information on user beliefs, RS designers can understand what items a user would consume \textit{without} recommendation and then predict how providing information to the user would change her beliefs and resulting consumption decisions. Thus,  our evaluation measure determines the value of a recommendation based on the marginal welfare gain associated with providing a user with a recommendation over what the user would do without it. We discuss how this provides an additional rationale as to why ``accurate" recommendations are not always good recommendations.

@@

The first set of related works studies the extent and implications of filter bubbles. \cite{pariser2011filter} first informally described the idea of the ``filter bubble'' which is that online personalization services would lead users down paths of increasingly narrower content so that they would effectively be isolated from a diversity of viewpoints or content. Following this a number of empirical studies, done in various disciplines, have since studied the extent to which this phenomenon exists in a wide range of contexts \cite{flaxman2016filter,hosanagar2013will,moller2018blame,nguyen2014exploring}. The most relevant to our study is \cite{nguyen2014exploring} who study whether this effect exists in the context of movie consumption. They find that even users whose consumption choices are not guided by recommendations exhibit behavior consistent with ``filter bubbles'' and that RSs can actually increase the diversity of the content that users consume. To our knowledge there are no theoretical models that rationalize these empirical findings and we provide a theoretical framework through which to view this problem. Furthermore, we provide a clear mechanism that drives such effects and how recommendation interacts with them.

@@

Another set of papers has examined whether RSs can lead users to become increasingly homogenized. \cite{celma2008hits, treviranus2009value} show that incorporating content popularity into RSs can lead to increased user homogenization. \cite{chaney2018algorithmic} shows how user homogenization may arise from training RSs on data from users exposed to algorithmic recommendations. \cite{fleder2009blockbuster} show that homogenization can increase due to a popularity recommendation bias that arises from lack of information about items with limited consumption histories. We show similar results as previous work where RSs lead to increased user homogenization. However, the mechanisms behind this differ from existing work as homogenization arises due to the fact that recommendation leads users to coordinate their consumption decisions in certain portions of the product space.

@@

Another set of papers studies the impact of human decision-making on the design and evaluation of RSs. \cite{chen2013human} surveys the literature on the relationship between human decision making and RSs. The closest set of papers pointed out in this survey are those related to preference construction \cite{bettman1998constructive, lichtenstein2006construction} whereby users develop preferences over time through the context of a decision process. We point out that the true underlying preferences of users may be stable over time, but, due to the nature of items in contexts where RSs are deployed, they have incomplete information of their valuations and both consumption and recommendation provide them with information to reduce their uncertainty. Thus, the primary insight of our paper is that user beliefs and how users update their beliefs about similar items after consumption are important and previously unconsidered elements of human decision making that are critical for understanding the design and consequences of RSs. Within this literature, \cite{celma2008new, cremonesi2013user, pu2011user} focus on ``user-centric" approaches to recommendation whereby user evaluation of the usefulness of recommendation is a key evaluation measure. Our evaluation measure is similar, but, unlike previous approaches, emphasizes the importance of user beliefs. Finally, \cite{hodgson2019horse} considers a similar model as ours where users engage in ``spatial learning'' and exploit the correlation of their preferences in the environment, but consider it in the context of search for a single item.

@@

\xhdr{User Learning}
When a user consumes an item $n$ she learns the realized value for that item. We assume that learning about the value of item $n$ reveals more about the value associated to items that are closer to it, which captures the idea that trying an item provides more information about similar items than about dissimilar ones.\footnote{\cite{schulz2019structured} empirically studies how individuals solve sequential decision-making problems under uncertainty in large choice sets in the context of mobile food delivery orders. They find that individuals engage in similarity-based generalizations where learning about the realized value of a particular good provides them with information about similar items. We incorporate this finding into our model in a stylized manner.} In order to have a well-defined notion of similarity we need to define a distance function between items, which we define as $d(n,m):=\min\{ \lvert m - n \rvert ,N - \lvert m - n \rvert \}$ where $m$ and $n$ are indices of items in $\mathcal{J}$. This distance function is not intended to model the intricacies of a realistic product space, but instead to provide a stylized product space to help us understand the effects of informational spillovers on user behavior. The basic intuition, in the context of movie consumption, is that a user watching \textit{John Wick} (item $n$) provides more information about how much she will like \textit{John Wick: Chapter Two} (item $m$) than \textit{Titanic} (item $q$) since $d(n, m) < d(n, q)$. 
\par
We consider that the entry of $n$-th row and the ($m$)-th column of $\Sigma_i$ is given by $\sigma_i^2 \rho^{d(n,m)}$, and that of $\Sigma$ is given by $\sigma^2 \rho^{d(n,m)}$. The scalar $\rho \in [0,1]$ therefore impacts the covariance structure: a higher $\rho$ implies that learning the utility of $n$ is more informative about products nearby. Informativeness, for any $\rho \in (0,1)$, is decreasing in distance. The particular distance function that we utilize leads to this covariance structure being simple, where the $(n,n+1)$-th entry in the covariance matrix is $\sigma^{2} \rho$, the $(n,n+2)$-th entry is $\sigma^{2} \rho^2$, etc.\footnote{This exponential decay correlation structure can be related to the tenet of case-based similarity of \cite{gilboa1995case} -- see \cite{billot2008axiomatization} for an axiomatization of exponential similarity.}
\par
The precise updating rule is as follows. Recall that at time $t$ the user's consumption history is given by $C_{i}^{t}$ and we denote the utility realizations of these items as $c_t$. We denote $\mu_t$ as the initial mean beliefs the user has over the items in $C_{i}^{t}$ and $\mu_{N-t}$ as the initial mean beliefs the user has over the remaining $N-t$ items, $\mathcal{J} \setminus C_{i}^{t}$. We partition the covariance matrix as follows:
\[ \Sigma =  \left( \begin{array}{cc}
\Sigma_{(N-t, N-t)} & \Sigma_{(N-t,t)} \\
\Sigma_{(t,N-t)} & \Sigma_{(t,t)}
\end{array} \right)
\]
Thus, after consuming the items in $C_{i}^{t}$ the resulting beliefs over the remaining items are given by $\mathcal{N}(\bar{\mu}, \bar{\Sigma})$ where $\bar{\mu}$ and $\bar{\Sigma}$ are as follows:
\begin{align*}
\bar{\mu} \mid c_t = \mu_{N-t} + \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1}(c_t - \mu_t) \\
\bar{\Sigma} \mid c_t = \Sigma_{(N-t,N-t)} - \Sigma_{(N-t,t)} \Sigma_{(t,t)}^{-1} \Sigma_{(t,N-t)}
\end{align*}

@@

\xhdr{Recommendation.}
Our model of recommendation is stylized in order to provide qualitative insights into how recommendation shapes behavior, instead of focusing on the details of how RSs are implemented in practice. We model recommendation as giving users information about the valuation of the products.
\par

@@

We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, \ standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$. The range of considered parameter values cover the relevant portions of the parameter space in order to provide full insight into the behavior of the model. When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest. We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.\footnote{In the supplementary material we include the same exercises presented here for $N = 100$ and $N = 500$ and show that the results are qualitatively similar to those reported here.}
\par

\begin{figure}
\caption{Local Consumption and Correlation, $N = 200$}
\begin{subfigure}{.3\linewidth}
  \centering
  \includegraphics[width=1.0\linewidth]{figures/rho_pos_consumption_dist_N_200T_20_overall.pdf}
  \label{fig:sfig1}
\end{subfigure}%
\begin{subfigure}{.3\linewidth}
  \centering
  \includegraphics[width=1.0\linewidth]{figures/rho_zero_consumption_dist_N_200T_20_overall.pdf}
  \label{fig:sfig2}
\end{subfigure}
\caption*{\scriptsize Notes: The figure shows the consecutive consumption path difference between the no recommendation, recommendation, and oracle regime. The figure on the left displays the average consecutive consumption distance aggregating over simulations where $\rho > 0$ and the figure on the right displays the average consecutive consumption distance aggregating over simulations where $\rho = 0$.}
\label{fig:correlation_consumption_path}
\end{figure}

\begin{figure}[ht]
\caption{Relationship between Local Consumption and $\rho$, $N = 200$}
\begin{subfigure}{.3\textwidth}
\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20.pdf}
\end{subfigure}
\begin{subfigure}{.3\textwidth}
\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_partial.pdf}
\end{subfigure}
\begin{subfigure}{.3\textwidth}
\includegraphics[width=\linewidth]{figures/rho_consumption_dist_N_200T_20_omni.pdf}\\
\end{subfigure}
\begin{subfigure}{.3\textwidth}
\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20.pdf}
\end{subfigure}
\begin{subfigure}{.3\textwidth}
\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_partial.pdf}
\end{subfigure}
\begin{subfigure}{.3\textwidth}
\includegraphics[width=\linewidth]{figures/gamma_consumption_dist_N_200T_20_omni.pdf}
\end{subfigure}%
\caption*{\scriptsize Notes: Each figure plots the average consecutive consumption distance across time as the inherent correlation between the valuation of the items, $\rho$. The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime. Notes: Each figure plots the average consecutive consumption distance across time as the risk aversion level of users, $\gamma$, varies. The top left displays the no recommendation regime, the top right displays the recommendation regime, and the bottom displays the oracle regime.}
\label{fig:local_consumption_across_rho}
\end{figure}

\section{Results}
\subsection{Local Consumption and Filter Bubbles}
We characterize ``filter bubble'' effects in our model as the degree to which users engage in \textit{local consumption}. We define local consumption in terms of the average consumption distance between the items consumed by the users at time $t-1$ and $t$. Thus, in the context of our model, filter bubble effects arise when the average consumption distance decreases over time and, across regimes, when the levels are lower for a given recommendation regime compared to another.

@@

\subsection{User Welfare and Product Diversity}
In this section we primarily focus on the impact of recommendation on user welfare and the overall diversity of the items that they consume. While in the previous section we looked at the distance between consecutive items, in this section we focus on a diversity measure that considers the entire consumed set of items. The diversity measure we utilize is common in RS literature (e.g. \cite{ziegler2005improving}) which is the average normalized pairwise distance between the consumed items:
$$D_i:=\frac{1}{N}\frac{1}{T(T-1)}\sum_{n,m \in C_i^T: n \ne m} d(n,m)$$

\begin{finding}\label{finding_diversity}
The impact of recommendation on product diversity:
\begin{enumerate}
\item When $\rho = 0$, product diversity is the same across all three recommendation regimes;
\item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime. This effect is amplified as $\rho$ increases as well as when users become more risk-averse.
\end{enumerate}
\end{finding}
\par 
Finding~\ref{finding_diversity} summarizes the main results on product diversity. As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes. The over-exploitation of information spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime. As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases. In the supplementary material (Figure 5), we show how diversity varies as $\rho$ increases across the three regimes that we consider. There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in the supplementary material (Figure 9). The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
\par

\begin{figure}[ht]
\caption{Diversity vs. Welfare, \textbf{No Recommendation}, N = 200}
\begin{subfigure}{.3\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_n_200.pdf}
\end{subfigure}
\begin{subfigure}{.3\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_n_200.pdf}
\end{subfigure}\\
\begin{subfigure}{.3\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial_n_200.pdf}
\end{subfigure}
\begin{subfigure}{.3\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial_n_200.pdf}
\end{subfigure}\\
\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the no recommendation regime. The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$. Notes: The figure plots the relationship between diversity and welfare under the recommendation regime. The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
\end{figure}
\addtocounter{figure}{-1}
We now study how recommendation impacts user welfare in our model. In our model users make consumption decisions that maximize their current period ex-ante utility that depends on their beliefs in that period. Thus, from an ex-ante perspective, they make optimal decisions, but our primary interest is in understanding how the ex-post, or realized, utility varies across regimes and parameter values. We define user's \textit{ex-post} welfare as the average of the realized values, controlling for the effect of $T$:
$$W_i:= \frac{1}{T}\sum_{n \in C_i^T} x_{i,n}$$
